A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-02-04 analysis/paper/paper.Rmd
edit based on BF's comments, stop at model specification

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationship between entities. Different from traditional static networks, statistical approaches to network analysis, such as exponential random graph models (ERGMs), provides a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!--burial study- network analysis-->
Burial analysis by archaeologists is an approach to understand past social structures through the study of the physical traces of mortuary practices. The material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. By studying burial forms, grave goods, and ritual behaviors that structure the material configuration of burials, we can infer social status and relations between members of a community that give an insight to social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theory, the recent development of social network analysis provides many new tools to visualize and analyze relationships in archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation [@Brandes2013; @Collar2015]. A network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This paper will use a novel Bayesian approach on exponential random graph models for networks to investigate the formation of relationships between burials in northeastern Taiwan to explore the indirect impact of foreign presence on social organization.

@@

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show the evidence of accumulation patterns of imported goods, such as ornaments and porcelains, in residential areas and burial contexts around the same time as European contacts in the 17th century [@Chen2007; @LWandBM2020ornament]. Imported goods were treated as prestige or wealth in those Indigenous societies that were commonly mentioned in Spanish historical records in the 17th century (1624-1642) [@Borao2009; @LiandWu2006]. For example, a document written by a Spanish priest described a Spanish soldier using carnelian beads to exchange for natural resources with local Indigenous people, because of their high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status by their community [@LiandWu2006]. Despite the fact that we recorded an uneven distribution of imported goods in burials, the degree of differentiation across burials over time remains unclear. The introduction of foreign prestige goods to local Indigenous communities might have occurred before European presence became established, but was maintained and amplified after European arrival due to local competition for power [@Borao2009; @LiandWu2006; @LWandBM2020ornament; @Ueda2016]. A general model to summarize this situation in northeastern Taiwan might be that the influence of a colonial power, combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. The observed uneven distribution in Northeastern Taiwan may be explained as a result of unequal access to trade networks when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

<!-- archaeological specific case study -->
In this study, we explore social changes in a pericolonial context at Kiwulan (Figure \@ref(fig:figure-KWL-map)), an Iron Age site in northeastern Taiwan, which was occupied from 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) did European colonial activities in 17th century Taiwan result in an increased social inequality in an Indigenous society in ways that can be detected by analysis of burial networks? (2) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of statistical modeling for burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand reactions to indirect impacts of European colonization.

@@

<!-- ERGM in archaeology, and limitations --> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. Although their models do not fit perfectly with the observed network of archaeological data, they proposed that ERGM is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. However, those studies also point out some limitations of ERGM, such as sensitivity to missing data and less ability to handle uncertainty. In addition, it is difficult in ERGM to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs in a Bayesian framework.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, where the acceptance probability is tractable that give a better convergence result. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. After fitting an ERMG with an MCMC algorithm, the Bayesian approach generates posterior probabilities incorporating our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by looking at the posterior mean and 95% credible intervals, or the odds ratios (conditional probability of observing an edge) that can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting an significant effect of a parameter is that its credible interval does not contain zero, or an odds ratio is greater/less than one, meaning a significant positive/negative effect [@Harris2013, pp. 51]. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data is missing.

@@

<!-- archaeological setting -->
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes for the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) [@Chen2007]. Excavations revealed abundant trade goods, including imported ceramics and stonewares, metal artifacts, glass beads, and stone beads, indicative of frequent trade activities [@Chen2007]. A total of 90 burials were unearthed that are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and _in-situ_ wooden posts, suggesting a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that have not yet been well studied.

@@

<!-- chronology -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [@LWandBM2020ornament; @Chen2007]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. There are `r disturb` burials heavily disturbed by modern construction that are also excluded because we cannot determine their chronology. Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, and fish-shaped ornaments. Previous studies indicate that these items are considered as high-value across different archaeological contexts [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. Historical accounts written by Europeans and the Han Chinese also support these items as prestige goods [@Borao2009; @LiandWu2006].

@@

<!-- our hypothesis --> 
Our pericolonial impact hypothesis is that rare, high quality burial goods accumulated in only a few individuals' burials after the European presence because of increased social inequality at Kiwulan due to differential access to these goods. A related effect of the European presence in northeastern Taiwan may have been manipulation of the European colonial image by ambitious local Indigenous individuals for building their personal status and power, recognized as network strategies to achieve power [@Drennan2010; @Feinman2000]. If social inequality gradually increased in Kiwulan as we hypothesize, then we expect to observe a network with higher degree (popularity) and less transitivity (cohesion), along with greater differences in wealth accumulation according to burial goods. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- burial data and the definition of tie formation  --> 
We used burial goods in common to construct ties for representing a flow of goods between burials. Since trade beads are commonly found across burials with substantial differences in quantities, we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting based on historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three levels, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculated by the total number of a prestige item from burial context over the number of prestige items in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship. The networks constructed based on these principles show that the network after the European presence has higher node degrees in general (Figure \@ref(fig:figure-network-diagrams)).

```{r figure-network-diagrams, fig.cap = "A: Burial network before the European arrival, B: Burial network after the European arrival. The size of the nodes is proportional to the node degrees that means the number of connections to a node. The thickness of tie represents the number of goods in common between nodes."}
knitr::include_graphics(here::here("analysis", "figures", "005-network-diagrams.png"))
```

>>>>

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationships between entities. Statistical approaches to network analysis, such as exponential random graph models (ERGMs), provide a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!--burial study- network analysis-->
Burial analysis by archaeologists is an approach to understand past social structures through the study of the physical traces of mortuary practices. The material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. By studying burial forms, grave goods, and ritual behaviors that structure the material configuration of burials, we can infer social status and relations between members of a community that give an insight to social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theory, the recent development of social network analysis provides many new tools to visualize and analyze relationships in archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation [@Brandes2013; @Collar2015]. A network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. Burials with the same kinds of prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This paper will use a novel Bayesian approach on exponential random graph models for networks to investigate the formation of relationships between burials in northeastern Taiwan to explore the indirect impact of foreign presence on social organization.

@@

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show the evidence of accumulation patterns of imported goods, such as ornaments and porcelains, in residential areas and burial contexts around the same time as European contacts in the 17th century [@Chen2007; @LWandBM2020ornament]. Imported goods were treated as prestige or wealth in those Indigenous societies that were commonly mentioned in Spanish historical records in the 17th century (1624-1642) [@Borao2009; @LiandWu2006]. For example, a document written by a Spanish priest described a Spanish soldier using carnelian beads to exchange for natural resources with local Indigenous people, because of their high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status in their community [@LiandWu2006]. Despite the fact that we recorded an uneven distribution of imported goods in burials, the degree of differentiation across burials over time remained unclear. The introduction of foreign prestige goods to local Indigenous communities might have occurred before European presence became established, but was maintained and amplified after European arrival due to local competition for power [@Borao2009; @LiandWu2006; @LWandBM2020ornament; @Ueda2016]. A general model to summarize this situation in northeastern Taiwan might assume that the influence of a colonial power, combined with high local values attached to imported goods, might lead to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. The observed uneven distribution in Northeastern Taiwan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

<!-- archaeological specific case study -->
In what follows, we explore social changes in a pericolonial context at Kiwulan (Figure \@ref(fig:figure-KWL-map)), an Iron Age site in northeastern Taiwan, which was occupied from 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. We argue that the involvement in the long-distance trade stimulated by the Europeans might lead to increased social inequality at Kiwulan. This argument would be supported by changes in the structure of networks built on burial data based on the assumption that burials can reflect social structure. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival in 17th century Taiwan. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) what are the major variables affecting or forming the unequal social positions represented by higher degree of network popularity (e.g., a few nodes have more relationships with other nodes) that might hint at social heterogeneity influence by the European arrival? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in an indirect way by examining burial data. We also introduces a novel Bayesian approach on statistical modeling for networks that is suitable for networks with smaller sizes.

@@

<!-- ERGM in archaeology, and limitations --> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. Although their models do not fit perfectly with the observed network of archaeological data, they proposed that ERGM is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. However, those studies also point out some limitations of ERGM, such as sensitivity to missing data and less ability to handle uncertainty. In addition, it is difficult in ERGM to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with a larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue where probability models tend to overestimate a small number of extreme graphs by assigning too much weight, e.g., in extreme cases such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs by means of a Bayesian framework.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. With the exchange algorithm, Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives better convergence results. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with an MCMC algorithm, a Bayesian approach generates posterior probabilities that incorporate our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by examining the posterior mean and 95% credible intervals, or the odds ratios (conditional probability of observing an edge) that can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting an significant effect of a parameter is that its credible interval does not contain zero, or an odds ratio is greater/less than one, meaning a significant positive/negative effect [@Harris2013, pp. 51]. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data is missing.

@@

<!-- archaeological setting -->
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes for the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) [@Chen2007]. Excavations revealed abundant trade goods, including imported ceramics and stonewares, metal artifacts, glass beads, and stone beads, indicative of frequent trade activities [@Chen2007]. A total of 90 burials were unearthed that are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with rare goods, also prestige goods in local culture, were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that have not yet been well studied.

@@

<!-- chronology -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [@LWandBM2020ornament; @Chen2007]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. There are `r disturb` burials heavily disturbed by modern construction that are also excluded because we cannot determine their chronology. Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, and fish-shaped ornaments. Previous studies indicate that these items are considered as high-value across different archaeological contexts based on their rarity and descriptions in historical records [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. Historical accounts written by Europeans mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006].

@@

<!-- our hypothesis --> 
Our "pericolonial impact hypothesis" is that rare, high quality burial goods accumulated in only a few individuals' burials after the European presence because of increased social inequality at Kiwulan due to differential access to these goods. A related effect of the European presence in northeastern Taiwan may have been manipulation of the European colonial image by ambitious local Indigenous individuals for building their personal status and power, recognized as network strategies to achieve power [@Drennan2010; @Feinman2000]. If social inequality gradually increased in Kiwulan as we hypothesize, then we expect to observe a network with higher degree (popularity) and less transitivity (cohesion), along with greater differences in wealth accumulation according to burial goods. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- burial data and the definition of tie formation  --> 
We used burial goods in common to construct ties for representing a flow of goods between burials. Since trade beads are commonly found across burials with substantial differences in quantities, we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting based on historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three ranks, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculated by the total number of prestige item from burial context over the number of prestige items in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

```{r figure-network-diagrams, fig.cap = "A: Burial network before the European arrival, B: Burial network after the European arrival. The size of each node is proportional to "node degrees" or the number of connections to a node. The thickness of tie represents the number of goods in common between two nodes."}
knitr::include_graphics(here::here("analysis", "figures", "005-network-diagrams.png"))
```
